On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-16 manuscripts/PNAS/chazthm_pnas.tex
chaz

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$.

In this note, we prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while significantly reducing the known sample complexity. Our results justify the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, as in \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research. The main motivation for this work, however, was to understand how seemingly universal representations emerge from various sparse coding models fit to natural data.

What we have shown here is that the sparse linear coding model generally produces a \textit{well-posed} inverse problem to be approximately solved by a numerical algorithm. This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In other words, a small perturbation of the data should result in only slightly different inferred parameters. In this regard, for the sparse linear coding model we demonstrate a linear relationship, \eqref{epsdel}, between measurement noise and the deviation of any solution from the true parameters, with explicit constants expressed in terms of these parameters. Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. We remark that our constants have been derived for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be smaller when sampled from a distribution.

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a sample size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. A reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this can in fact be achieved if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

@@

In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that some of the reproduce response properties of cortical neurons. Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early idea in neural theory \cite{pitts1947}. %Furthermore, they justify the hypothesis of \cite{Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).
Deep learning noise issues: \cite{goodfellow2014explaining} and neuroscience theory \cite{ganguli2012compressed}.

>>>>

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$. 
In the last several decades, from JPEG's dominance in image compression to the discovery of intrinsic "Gabor" wavelets in natural images to the detection of forgeries to artificial "cat detecting neurons" in machine intelligence, sparse linear coding has become a standard tool in signal analysis, allowing for the unsupervised discovery of myriad insights about natural signals. A compelling feature of the model is its simple instantiation of the principle of Occam's razor. For instance, a natural video can be modeled, quite accurately, as a linear combination of a small number of spatiotemporal "basis functions", each representing an archetypical feature latent in the data. In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that reproduce response properties of certain cortical neurons. The main motivation for this work was to understand how such seemingly universal representations emerge from sparse coding models fit to natural signals.

We have proven the uniqueness and stability of solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem}, conclusively demonstrating that this parsimonious model of (noisy) natural data is well-posed in the sense of Hadamard \cite{Hadamard1902}, a long-standing necessary property of reductionist models in science. This early concept can be paraphrased as the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In other words, a small perturbation of the data should result in only slightly different inferred parameters. In this regard, for the sparse linear coding model we demonstrate a linear relationship, \eqref{epsdel}, between measurement noise and the deviation of any solution from the true parameters, with explicit constants expressed in terms of these parameters. Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. 
%This work justifies the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, as in \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research.

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. 
%For instance, whether there is a sample size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. 
A reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this result enables a practical amount of data to determine the model if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

@@

We remark that our constants have been derived for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be smaller when sampled from a distribution.

 Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early idea in neural theory \cite{pitts1947}. %Furthermore, they justify the hypothesis of \cite{Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).
Deep learning noise issues: \cite{goodfellow2014explaining} and neuroscience theory \cite{ganguli2012compressed}.

Sparse coding also forms the theoretical basis of one of the few hypothesized methods of bottleneck communication between sparsely active neural populations.
